\documentclass[handout,space,nooutcomes]{ximera} 

\input{../preamble.tex}

\title{DeMoivre's Theorem}
\author{Brad Findell}
\begin{document}
\begin{abstract}
Problems exploring DeMoivre's Theorem.   
\end{abstract}
\maketitle



%The notation cis theta is shorthand for  

Thinking of the complex plane as ordered pairs of real numbers, the number $a+bi$ can be represented as the point $(a,b)$ or as a vector from the origin to
$(a,b)$.  If $a+bi = r(\cos\theta +i\sin\theta)$, for real numbers $a$, $b$, $r$, and $\theta$, then we say that the complex number has rectangular coordinates $(a,b)$ and polar coordinates $(r;\theta)$. We call $r$ the magnitude and $\theta$ the angle (or argument) of the complex number. Practice converting among these forms, particularly for special angles.

\begin{problem}
You have seen Euler's formula: $e^{i\theta}=\cos\theta +i\sin\theta$.  For a derivation of this formula, write the Taylor series for $e^{x}$, substitute $x=i\theta$, and organize the result into Taylor series for each of the real and imaginary parts.  What do you notice?  
% $\cos x$, and $\sin x$ by separating the real and imaginary parts of $e^{i\theta}$.  

\end{problem}

\begin{problem}
Let $z = \cos\theta +i\sin\theta$.  Where must $z$ lie on the complex plane for any $\theta$?
%Note that because 
%$\cos^2\theta+\sin^2\theta=1$, the complex number $\cos\theta +i\sin\theta$ lies on the 
%unit circle in the complex plane.  
%More generally, any complex number $a+bi$ can be expressed as $r(\cos\theta +i\sin\theta)$, with $r\ge 0$.  
\end{problem}

%\begin{problem}
%Use the angle addition identity to prove DeMoivre's theorem:  .
%\end{problem}

\begin{problem}
Suppose $z_1=r_1 e^{i\alpha}$ and $z_2=r_2 e^{i\beta}$, where $r_1, r_2, \alpha$, and $\beta$ are real numbers.  

With Euler's formula, we can write 
\begin{align*}
z_1& =r_1 e^{i\alpha}= r_1 (\cos\alpha + i\sin\alpha) \text{ and } \\
z_2& =r_2 e^{i\beta}=r_2(\cos\beta + i\sin\beta),  \text{ so } \\
z_1 z_2 & = r_1(\cos\alpha + i\sin\alpha)r_2(\cos\beta + i\sin\beta).
\end{align*}

But this is difficult to interpret.  

On the other hand, from the algebra of exponents, it follows that 
\[
z_1 z_2 = r_1e^{i\alpha}\cdot r_2 e^{i\beta}= r_1 r_2 e^{i(\alpha + \beta)}.  
\]

%\[
%z_1 z_2 = r_1(\cos\alpha + i\sin\alpha)r_2(\cos\beta + i\sin\beta)
%=r_1 r_2\left(\cos(\alpha+\beta) + i\sin(\alpha+\beta)\right). 
%\]

What these equations tell us about the geometry of complex multiplication?  
\begin{freeResponse}
\begin{hint}
When multiplying complex numbers, simply multiply their magnitudes and add their angles (arguments).  
\end{hint}
\end{freeResponse}
\end{problem}

\begin{problem}
Multiply $a + bi$ times $c + di$ for the special case when $a=\cos\alpha$, $b=\sin\alpha$, 
$c=\cos\beta$, $d=\sin\beta$.  Explain how this idea could help you reconstruct certain trigonometry formulas, if you had forgotten them.  

Explain, alternatively, how you could use your knowledge of trigonometry to prove the angle-addition property of complex multiplication.  
\end{problem}

%
%\begin{problem}
%Prove the zero product property for complex numbers.  Give an example of a number system in which the zero product property does not hold.
%\vfill 
%\end{problem}
%
Let's step back for a moment to see whether we can generate the same understanding about complex multiplication without Euler's formula---and without trigonometry.  
\begin{problem}
Prove that multiplying by a complex number $w = a + bi$ is a composition of a rotation and a dilation and specify both the angle of rotation and the dilation factor.  Hints: For a generic complex number $z$, how do $za$ and $zbi$ relate to $z$?  How does $zw$ relate to $za$ and $zbi$?   Here is GeoGebra sketch that might help: 
\url{https://www.geogebra.org/m/f3dwskkd}.  
\end{problem}

\begin{problem}
Prove:  $(\cos\theta + i\sin\theta)^2=\cos 2\theta + i\sin 2\theta$.  Describe the formula in words.    
\end{problem}

\begin{problem}
Prove DeMoivre's Theorem:  $(\cos\theta + i\sin\theta)^n=\cos n\theta + i\sin n\theta$, for $n=1, 2, \dots$.  Hint: Use mathematical induction on $n$. 
\end{problem}

\subsection*{Finding roots}
DeMoivre's theorem is often used to find roots of complex numbers.  
To derive the formula, observe first that, in polar form, complex numbers are not uniquely identified.  In particular, 
\[
\cos\theta + i\sin\theta =\cos (\theta+2k\pi) + i\sin (\theta+2k\pi), 
\]
for any integer $k$. Thus, by DeMoivre's theorem:  
\begin{align*}
\left[ \cos \left(\frac{\theta+2k\pi}{n}\right) 
+ i\sin \left(\frac{\theta+2k\pi}{n}\right) \right]^n 
&= \cos (\theta+2k\pi) + i\sin (\theta+2k\pi) \\
&= (\cos\theta + i\sin\theta), 
\end{align*}
which means that 
\[
\cos \left(\frac{\theta+2k\pi}{n}\right) 
+ i\sin \left(\frac{\theta+2k\pi}{n}\right)
\]
is an $n^\text{th}$ root of $(\cos\theta + i\sin\theta)$ for any integer $k$.  To ensure that these $n^\text{th}$ roots are distinct, we typically choose $k= 0, 1, \dots, n-1$.  This approach works for any complex number on the unit circle.  

Now we generalize this approach to find the $n^\text{th}$ roots any non-zero complex number $z=r(\cos\theta + i\sin\theta)$, where $r>0$.  Because $r$ is a positive real number, it has exactly one positive real $n^\text{th}$ root, called its ``principal $n^\text{th}$ root,'' which we can then unambiguously denote as $\sqrt[n]{r}$ or as $r^{1/n}$.  Then the $n$ distinct $n^\text{th}$ roots of $z$ are as follows: 
\[
\sqrt[n]{r}\left[\cos \left(\frac{\theta+2k\pi}{n}\right) 
+ i\sin \left(\frac{\theta+2k\pi}{n}\right)\right], \text{ for } k= 0, 1, \dots, n-1. 
\]
Notice that this means that the $n$ roots are neatly arranged on a circle of radius 
$\sqrt[n]{r}$, separated by a uniform angle of $\frac{2\pi}{n}$ radians.  

\begin{problem}
Use DeMoivre's Theorem to find: 
\begin{enumerate}
\item The $5^\text{th}$ roots 1.  
\item The $4^\text{th}$ roots $i$.  
\item The $3^\text{rd}$ roots $-8i$.  
\end{enumerate}
Note: You may express your answers in either radians or degrees.  
\end{problem}



\end{document}
