\documentclass[space,nooutcomes]{ximera} 

\input{../preamble.tex}

\title{Fifth Roots of Unity}
\author{Brad Findell}
\begin{document}
\begin{abstract}
A sequence of problems deriving the fifth roots of unity.   
\end{abstract}
\maketitle



\begin{problem}

To find the fifth roots of unity, we need to solve the equation: 
\begin{equation}
x^5 = 1
\end{equation}
Or equivalently, 
\begin{equation} \label{eqA}
\answer{x^5 - 1} = 0
\end{equation}

\begin{problem}
Clearly $x=1$ is a solution, so $\answer{(x-1)}$ must be a factor.  

\begin{problem}
The left side of (\ref{eqA}) factors as follows: 
\begin{equation}
(x-1)\left(\answer{x^4+x^3+x^2+x+1}\right)=0
\end{equation}
\end{problem}
\end{problem}
\end{problem}

\begin{problem}
From previous explorations, we know that roots of unity all lie on the $\answer[format=string]{unit circle}$ (two words). 

If $x$ is an $n^\text{th}$ root of unity, what do we know about these quantities: 
\begin{itemize}
\item $x^{-1}$ 
\item $\overline{x}$
\item $x+x^{-1}$
\end{itemize}

Answers:
\begin{enumerate}
\item $x^{-1}$ is an $n^\text{th}$ root of $\answer[format=string]{unity}$.  
\item $\overline{x} = \answer{x^{-1}}$. 
\item $x+x^{-1}$ is a $\answer[format=string]{real}$ number.  
\end{enumerate}
\begin{problem}
Explanations: 
\begin{enumerate}
\item Because $x^n=1$, it follows that 
\[
\left(x^{-1}\right)^n = \answer{x^{-n}} = \left(x^n\right)^{-1} = 1^{-1} = 1. 
\]
\item Suppose $x=a+bi$, with $a$ and $b$ real numbers. Because $x$ is on the unit circle, $a^2+b^2 = \answer{1}$.  Then 
\[
x^{-1}=\frac{1}{a+bi}=\frac{1}{a+bi}\cdot\frac{a-bi}{\answer{a-bi}}=\frac{a-bi}{\answer{a^2+b^2}}=\frac{a-bi}{1}=\overline{x}
\]
\item For any complex number, $x+\overline{x}$ is a $\answer[format=string]{real}$ number.  When $x$ is on the unit circle, 
\[
x + \overline{x} = x + x^{-1} = x + \frac{1}{x}\text{ is a }\answer[format=string]{real}\text{ number.}
\]
So let's organize the equation around $x + \frac{1}{x}$.  
\end{enumerate}
\end{problem}
\end{problem}


\begin{problem}

To solve: 
\begin{equation}
x^4+x^3+x^2+x+1=0,
\end{equation}

Divide by $x^2$, and then organize around powers of $x + \frac{1}{x}$: 
\begin{align}
x^2+x+1+\frac{1}{x}+\frac{1}{x^2}&=0 \\
\left(x^2+ \answer{2} + \frac{1}{x^2}\right)+ \left(x+\frac{1}{x}\right)+\answer{-1}&=0 \label{eqB}
\end{align}
\begin{hint}
The expression inside the first parentheses of (\ref{eqB}) needs to be equivalent to 
\[
\left(x+\frac{1}{x}\right)^2
\]
\end{hint}
\begin{problem}
Now let $y = x + \frac{1}{x}$, and the equation becomes: 
\begin{equation} \label{eqC}
\answer{y^2+y-1}=0
\end{equation}

Solve equation (\ref{eqC}) for $y$.

How many solutions?  $\answer{2}$ 

\begin{problem}
Enter the solutions to (\ref{eqC}), from least to greatest: 

\begin{prompt}
$y = \answer{\frac{-1-\sqrt{5}}{2}}$

$y = \answer{\frac{-1+\sqrt{5}}{2}}$
\end{prompt}

\end{problem}
\end{problem}
\end{problem}

\begin{problem}
We defined $y$ as follows: 
\begin{equation} \label{eqD}
y = x + \frac{1}{x}.  
\end{equation}
Solve equation (\ref{eqD}) for $x$ in terms of $y$.  
\begin{hint}
Multiply to clear the fractions, then recognize that the (\ref{eqD}) is quadratic in $x$.  
% $x^2 - yx + 1 = 0
\end{hint}

How many solutions?  $\answer{2}$ 
\begin{problem}
\begin{prompt}
Enter the solutions to (\ref{eqD}), for $x$ in terms of $y$, as a conjugate pair: 

$x = \answer{\frac{y}{2}} \pm \answer{\frac{\sqrt{y^2-4}}{2}}$

\end{prompt}

\end{problem}
\end{problem}


\begin{problem}
Now put the previous answers together to yield all $\answer{4}$ complex solutions $x$ to the equation
\begin{equation} \label{eqE}
x^4+x^3+x^2+x+1=0.
\end{equation}

Enter the solutions to (\ref{eqE}) as conjugate pairs: 

Quadrants 1 and 4: $\left(\answer{\frac{-1+\sqrt{5}}{4}}\right) 
     \pm \left(\answer{\frac{\sqrt{10+2\sqrt{5}}}{4}}\right)i$

Quadrants 2 and 3: $\left(\answer{\frac{-1-\sqrt{5}}{4}}\right)
     \pm \left(\answer{\frac{\sqrt{10-2\sqrt{5}}}{4}}\right)i$ 



\end{problem}


\end{document}
